### 2025-03-31 decision-making.qmd
20250331 - reinforcement

@@

### Reinforcement Schedules {#sec-reinforcementSchedules}

Reinforcement schedules are depicted in @fig-reinforcementSchedules, as adapted from @Spielman2020.

@@

>>>>

The National Football League (NFL) used to be adamantly against sports gambling.
They wanted to protect the integrity of the game.
They felt that sports gambling might lead to match fixing, point shaving, and other forms of corruption that would undermine the legitimacy of competition in the game.
Match fixing occurs when players, referees, coaches, or others conspire to manipulate the outcome of the game.
Point shaving involves manipulating the score—commonly by a player intentionally underperforming (even if it does not lead to "throwing the game" and causing the team to lose).
Match fixing and point shaving can occur for a variety of reasons, but a common reason is to ensure that particular gamblers win their bet.
Another reason the NFL was against sports betting was that, prior to the U.S. Supreme Court overturning the Professional and Amateur Sports Protection Act (PASPA) in 2018, there were legal restrictions on gambling that prevented sports betting in many parts of the country.

However, after the legalization of sports betting following the overturning of PASPA in 2018, the NFL has changed their tune and are now strongly in favor of sports betting.
The rise of legal sports betting has created massive revenue streams for the NFL through partnerships, sponsorships, and advertising deals with sportsbooks.

@@

### Reinforcement {#sec-reinforcement}

Reinforcement involves a (typically appetitive) stimulus that increases the frequency of a behavior.
For instance, reinforcers could include things like food, praise, attention, money, etc.
Thus, sports gambling can provide reinforcement through the receipt of money.

#### Continuous Reinforcement {#sec-reinforcementContinuous}

In general, when you want to train a new behavior, continuous reinforcement is the fastest way to do so.
Continuous reinforcement means rewarding the behavior each time that it occurs.
For instance, a parent might use continuous reinforcement to train their child to put their toys away—the parent may praise them or give them physical affection each time they put their toys away.
However, continuous reinforcement is susceptible to extinction.
Extinction means cessation of the target behavior (putting the toys away, in this example).
That is, for a previously continusouly reinforced behavior, if the behavior stops being rewarded, the person is less likely to continue the behavior.
Thus, to make a behavior less susceptible to extinction, [intermittent reinforcement](#sec-reinforcementIntermittent) may be used.

#### Intermittent Reinforcement {#sec-reinforcementIntermittent}

Intermittent reinforcement (also called partial reinforcement) means rewarding the behavior sometimes but not everytime.
There are various approaches to intermittent reinforcement, called schedules of reinforcement.
The four primary reinforcement schedules are: fixed interval, fixed ratio, variable interval, adn variable ratio.
The reinforcement schedules are depicted in @fig-reinforcementSchedules, as adapted from @Spielman2020.

@@

A *fixed interval* reinforcement schedule occurs when the person is rewarded after a set amount of time.
For instance, if the reward becomes available every 30 minutes, the person receives the reward the first time they engage in the behavior after the reward becomes available.
As an example, consider a surgery patient who can receive a painkiller after pressing a button, but only up to once per hour.

A *variable interval* reinforcement schedule occurs when the person is rewarded after a varying (and unpredictable) amount of time.
As an example, consider a restaurant manager who is paid a bonus if the food inspector, who comes at unpredictable times, gives the restaurant a good rating.

A *fixed ratio* reinforcement schedule occurs when the person is rewarded after a set number of responses.
As an example, consider a coffee shop that offers a free coffee after every 10 purchases that a person makes.

A *variable ratio* reinforcement schedule occurs when the person is rewarded after a varying (and unpredictable) number of responses.
As an example, consider a slot machine that provides a large monetary reward (paired with the machine lighting up and bells going off) after a varying number of attempts.
Many forms of gambling involve variable ratio reinforcement.

The variable ratio reinforcement schedule is the least susceptible to extinction.
People can go without reinforcement many times and may still continue to gamble out of the hope for a large reward on the next attempt.
The low susceptibility of variable ratio reinforcement to extinction is, in part, why gambling (which often involves variable ratio reinforcement) can be highly addictive despite people's tendency to lose money gambling.

### Does Fantasy Football Luck or Skill? {#sec-luckOrSkill}

Fantasy football (and sports betting, more generally) is not 100% luck.
Skill can be involved in identifying undervalued assets—whether in terms of stocks in the stock market or professional football players, where inside information can be especially valuable (and thus illegal to profit from).
To evaluate the percentage of variability in fantasy football that is attributable to luck versus skill, @Getty2018 evaluated the extent to which one's performance in the first half of the season was correlated with one's performance in the second half of the season, under the assumption that underlying skill would lead to persistence of performance across time.
The authors examined FanDuel daily fantasy sports (DFS) leagues, which allow the user to set an entirely new lineup each week from all players, based on salary constraints ( so a user cannot just select the best player at every position).
The authors estimated that ~55% of the variability in fantasy football performance across time was due to skill and 45% was due to luck [@Getty2018].
Thus, performance in fantasy football is around half and half luck versus skill.

Because fantasy football is not 100% luck, it is not the same as a slot machine.
However, there are still unpredictable elements (e.g., injuries), and there is considerable luck involved.
As a result, sports betting and gambling on fantasy football still involve components of [variable ratio reinforcement](#sec-reinforcementIntermittent) that lend them to being potentially highly addictive.
